State of the Practice for Lattice Boltzmann Method Software

### 2022-02-10 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Revisions up to and including Section 7: Comparison of Principles, Process and Methodologies to Research Software in General

Three of the items that appear in Table~\ref{artifactspresent} do not explicitly
appear in the software development guidelines: i) list of related publications,
ii) performance information and iii) video guides.  The occurrence of the first
two items in this list are likely more a consequence of how we collected our
list of LBM artifacts, rather than being a unique practice of LBM developers. We
highlighted publications and performance information because we were explicitly
looking for both as part of filling in our measurement template. Related
publications and performance information certainly appear in other research
software projects; the practices just aren't highlighted in guidelines because
they are implicit in other documentation recommendations, such as theory
documentation and user guides. The observation of a video guide artifact is more
novel.  Other projects likely use videos, but the practice has apparently not
yet made it into community guidelines.  Putting aside the debate of the
effectiveness of learning from print versus video, video documentation is
growing in popularity, with a majority of Generation Z learners preferring video
to print \citep{Genota2018}.

Although the LBM community participates in most of the practices we found listed
in the general research software guidelines, some recommended practices were not
observed or rarely observed. In addition to the items in the last column of
Table~\ref{artifactspresent}, we can add the following community recommended
items that we rarely, if ever, observed:

@@

Software tools are used to support the development, verification, maintenance,
and evolution of software, software processes, and artifacts \citep[p.\
501]{GhezziEtAl2003}. Many tools are used by LBM software packages, as
summarized in Table~\ref{tbl_tools}.  The tools are subdivided into development
tools, dependencies, and project management tools.  In this section we answers
\rqref{RQ_CompareToolsProjMngmnt} by comparing the tool usage in LBM to
their utilization in the research software community in general.

@@

Many of the software packages that were assessed were developed by teams of two
or more people. Their work needed to be coordinated and managed.
Table~\ref{tbl_tools} shows the types of project management tools that were
explicitly noted in the artifacts, web-pages, or interviews with the developers.
As with development tools and dependencies, it is possible that other types of
project management tools are used, but they were not visible in the artifacts to
which we had access.  Collaboration tools, most often email and video, are noted
as being used when developing the software projects. Project management software
was not explicitly mentioned, but it is possible that some of the projects use
such software. Many of the projects are located on GitHub, where the developers
use the platform to help manage their projects, especially bug related issues.
Most of the projects appear to use change tracking and version control tools.
They often use GitHub or GitLab for this. One package (SunlightLB) uses CVS.
Document generation tools are mentioned in 12 of the 24 projects. The tools
Sphinx and Doxygen are explicitly used in this capacity.

The poor adoption of version control tools that Wilson lamented in 2006
\citep{Wilson2006} has greatly improved in the intervening years.  From
Section~\ref{Sec_Maintainability} 67\% of LBM packages use version control
(GitHub, GitLab and CVS).  The proliferation of version control tools for LBM
matches the increase in the broader research software community.  A little over
10 years ago (in 2010) version control was estimated to be used in only 50\% of
research software projects \citep{Nguyen-HoanEtAl2010}, but even at that time
\citet{Nguyen-HoanEtAl2010} noted an increase from previous usage levels.  A
survey in 2018 shows 81\% of developers use a version control system
\citep{AlNoamanyAndBorghi2018}.  This matches with \citet{Smith2018} shows
active projects in mesh generation, geographic information systems and
statistical software for psychiatry increasing from 75\%, 89\% and 17\%
(respectively) to 100\%, 95\% and 100\% (respectively) over a four year period
ending in 2018.  (For completeness the same study showed a decrease in version
control tool usage for seismology software over the same time period (from 41\%
down to 36\%).  Almost every software guide cited in
Section~\ref{Sec_CompareArtifacts} mentions the advice to use version control.

Section \ref{SecSurfCorrectAndVerifiab} shows that continuous integration is
rare in LBM.  Recommended by many \citep{vanGompelEtAl2016} etc.

@@

Answers \rqref{RQ_CompareMethodologies}.  Most of the software packages do not
explicitly state in their artifacts the motivations or design principles that
were considered when developing the software. One package, Sailfish, indicates
in its artifacts that shortening the development time was considered in early
stages of design, with the developers opting for using Python and CUDA/OpenCL to
achieve this without sacrificing any computational performance. The goals of
that project are explicitly listed as performance, scalability, agility and
extendability, maintenance, and ease of use. The project scored well in these
categories during our assessment.

Processes, like methods, are ways of doing things, especially in an orderly way;
while methodologies are defined as systems of methods
\citep{ghezzi1991fundamentals}. It is not explicitly indicated in the artifacts
of most of the packages that development involved following any specific model,
like a waterfall or agile development model. One developer (ESPResSo) noted that
while no formal model is used, their development model is something similar to a
combination of agile and waterfall development models. The developer teams of
the LBM packages are fairly small, so it is feasible for them to be organized
without the need for such models.  \wss{some repetition from the previous
section - should try to figure out how to remove the repetition.}

Seven of the software packages contain artifacts recording the general
development process and status. Eleven of the packages explicitly convey that
they would accept outside contributors, but generally the teams are centralized,
often working at the same institution.

The developers that were interviewed all noted similar project management
processes. In teams of only a couple of developers, additions of new features or
major changes are discussed with the entire team. Projects with more than a
couple developers have lead developer roles. These lead developers review
potential additions to the software. One of the developers (ESPResSo) that was
interviewed noted that an ad hoc peer review process is used to assess major
changes and additions.  Peer review (also known as code review and technical
review) is a frequently recommended component of developing research software
\citep{HerouxEtAl2008, Givler2020, OrvizEtAl2017, USGS2019}.

Thirteen (54\%) of the 24 software packages use GitHub for managing the project,
including nine (60\%) of the 15 alive packages, and four of the nine (44\%) dead
packages. Two projects (Palabos, WaLBerla) use GitLab. This could be indicative
of a transition to such software development and version control tools for SCS.
Typically there are several simultaneous development branches in these projects.

Documentation was also noted as playing a significant role in the development
process, specifically with on-boarding new developers. A goal of documentation
is to lower the entry barrier for these new contributors. The documentation
provides information on how to get started, orients the user to artifacts and
the source code, and explains how the system works, including the so-called
simulation engine and interface. The use of document generation tools is
mentioned in the artifacts of 13 software packages, and was noted during
interviews with developers. Sphinx and Doxygen are the tools that were
mentioned.

Two types of software changes were discussed during interviews with developers.
One is feature additions, which arise from a scientific or functional need.
These changes involve formal discussions within the development team, and lead
developer participation is mandatory. The other change type is code refactoring,
which only sometimes involves formal discussions with the development team. New
developers were noted to play an increased role in these changes compared to the
former changes. Software bugs are typically addressed in a similar fashion as
code refactoring, and issue tracking is commonly used to manage these changes.

Interviews with the developers of software packages also revealed a more
frequent use of both unit testing and continuous integration in the development
process than was found by only assessing the artifacts. The use of automatic
installation processes is also common. Most often this involved a Make script.

@@

>>>>

Three of the items that appear in Table~\ref{artifactspresent} do not explicitly
appear in the software development guidelines: i) list of related publications,
ii) performance information and iii) video guides.  The occurrence of the first
two items in this list are likely more a consequence of how we collected our
list of LBM artifacts, rather than being a unique practice of LBM developers. We
highlighted publications and performance information because we were explicitly
looking for both as part of filling in our measurement template. Related
publications and performance information certainly appear in other research
software projects; the practices just aren't highlighted in guidelines because
they are implicit in other documentation recommendations, such as theory
documentation and user guides. The observation of a video guide artifact is more
novel.  \citet{Fogel2005} recommends videos for open source projects, and other
projects likely use them, but the practice has apparently not yet made it into
research software community's guidelines.  Putting aside the debate of the
effectiveness of learning from print versus video, video documentation is
growing in popularity, with a majority of Generation Z learners preferring video
to print \citep{Genota2018}.

Although the LBM community participates in most of the practices we found listed
in the general research software guidelines, some recommended practices were not
observed or rarely observed. For instance API documentation was rarely observed
for LBM software, but it is a frequently recommended artifact
\citep{SmithAndRoscoe2018, ThielEtAl2020, vanGompelEtAl2016, OrvizEtAl2017,
SSI2022, Zadka2018}.  In addition to the items in the last column of
Table~\ref{artifactspresent}, we can add the following community recommended
items that we rarely, if ever, observed:

@@

Software tools are used to support the development, verification, maintenance,
and evolution of software, software processes, and artifacts \citep[p.\
501]{GhezziEtAl2003}. Many tools are used by LBM software packages, as
summarized in Table~\ref{tbl_tools}.  The tools are subdivided into development
tools, dependencies, and project management tools.  In this section we answer
\rqref{RQ_CompareToolsProjMngmnt} by comparing aspects of the tool usage in LBM
to their utilization in the research software community in general.

@@

Many of the software packages that were assessed were developed by teams of two
or more people. Their work needed to be coordinated and managed.
Table~\ref{tbl_tools} shows the types of project management tools that were
explicitly noted in the artifacts, web-pages, or interviews with the developers.
As with development tools and dependencies, it is possible that other types of
project management tools are used, but they were not visible in the artifacts to
which we had access.  Collaboration tools, most often email and video, are noted
as being used when developing the software projects. Project management software
was not explicitly mentioned, but it is possible that some of the projects use
such software. Many of the projects are located on GitHub, where the developers
use the platform to help manage their projects, especially bug related issues.
Most of the projects appear to use change tracking and version control tools.
Document generation tools are mentioned in 12 of the 24 projects. The tools
Sphinx and Doxygen are explicitly used in this capacity.

The poor adoption of version control tools that Wilson lamented in 2006
\citep{Wilson2006} has greatly improved in the intervening years.  From
Section~\ref{Sec_Maintainability}, 67\% of LBM packages use version control
(GitHub, GitLab or CVS).  (From Table~\ref{gitrepodata}, 11/15 or 73\% of alive
packages use version control.)  The proliferation of version control tools for
LBM matches the increase in the broader research software community.  A little
over 10 years ago version control was estimated to be used in only 50\% of
research software projects \citep{Nguyen-HoanEtAl2010}, but even at that time
\citet{Nguyen-HoanEtAl2010} noted an increase from previous usage levels. A
survey in 2018 shows 81\% of developers use a version control system
\citep{AlNoamanyAndBorghi2018}.  \citet{Smith2018} has similar results, showing
version control usage for alive projects in mesh generation, geographic
information systems and statistical software for psychiatry increasing from
75\%, 89\% and 17\% (respectively) to 100\%, 95\% and 100\% (respectively) over
a four year period ending in 2018.  (For completeness the same study showed a
decrease in version control usage for seismology software over the same time
period, from 41\% down to 36\%).  Almost every software guide cited in
Section~\ref{Sec_CompareArtifacts} includes the advice to use version control.
The high usage of version control tools in LBM software matches the trend in
research software in general.

As mentioned in Section~\ref{SecSurfCorrectAndVerifiab}, continuous integration
is rarely used in LBM (3 of 24 packages or 12.5\%). This contrasts with the
frequency with which continuous integration is recommended in research software
development guidelines \citep{BrettEtAl2021, Brown2015, ThielEtAl2020,
Zadka2018, vanGompelEtAl2016}.  We could not find published data on the
frequency with which continuous integration is used in general research
software.  Our impression is that like LBM software, other research software
packages lag behind the recommended best practice of employing continuous
integration tools.

Interviews with the developers revealed a potentially more frequent use of both
unit testing and continuous integration, compared to what was observed from
studying the repository artifacts.

@@

This section answers research question \rqref{RQ_CompareMethodologies} by
comparing the principles, processes and methodologies used for LBM software to
what can be gleaned from the literature on research software in general. In our
data collection for LBM software, the software development process is not
explicitly indicated in the artifacts for most of the packages. However, during
our interviews one developer (ESPResSo) told us their non-rigorous development
model is like a combination of agile and waterfall. Employing a loosely defined
process makes sense for LBM software, given that the teams are generally small
and self-contained.  Although eleven of the packages explicitly convey that they
would accept outside contributors, generally the teams are centralized, often
working at the same institution.  Working at the same institution means that an
informal process can show success, since informal conversations are relatively
easy to have.

Interviews with developers confirmed a similar project management processes. In
teams of only a couple of developers, additions of new features or major changes
are discussed with the entire team. Projects with more than a couple developers
have lead developer roles. These lead developers review potential additions to
the software. One of the developers (ESPResSo) that was interviewed noted that
an ad hoc peer review process is used to assess major changes and additions.
Using peer review (also called technical review) matches with recommended
practice for research software \citep{HerouxEtAl2008, Givler2020, OrvizEtAl2017,
USGS2019}.

Two types of software changes were discussed during interviews with developers.
One is feature additions, which arise from a scientific or functional need.
These changes involve formal discussions within the development team, and lead
developer participation is mandatory. The other change type is code refactoring,
which only sometimes involves formal discussions with the development team. New
developers were noted to play an increased role in these changes compared to the
former changes. Software bugs are typically addressed in a similar fashion as
code refactoring.  Issue tracking is commonly used to manage these changes.

Our observations of a informally defined process, with elements of agile
methods, matches what has been observed for research software in general.
Scientific developers naturally use an agile philosophy \citep{AckroydEtAl2008,
CarverEtAl2007, EasterbrookAndJohns2009, Segal2005, HeatonAndCarver2015}, or an
amethododical process \citep{Kelly2013}, or a knowledge acquisition driven
process \citep{Kelly2015}.

Most of the software packages do not explicitly state the motivations or design
principles that were considered when developing the software. One package,
Sailfish, indicates in its artifacts that shortening the development time was
considered in early stages of design, with the developers using Python and
CUDA/OpenCL to achieve this without sacrificing performance. The Sailfish goals
are explicitly listed as performance, scalability, agility and extendability,
maintenance, and ease of use. The project scored well in these categories during
our assessment.  The quality priorities for Sailfish roughly match the
priorities observed for research software in general.
\citet{Nguyen-HoanEtAl2010} surveyed developers to find the following list of
qualities, in decreasing order of importance: reliability, functionality,
maintainability, availability, performance, flexibility, testability, usability,
reusability, traceability, and portability. The Sailfish list does not list
reliability or functionality, but we can safely assume those are implicitly high
priorities for any scientific project.  In earlier studies
\citet{KellyAndSanders2008} and \citet{CarverEtAl2007} both highlight how
important correctness is for research software.

During our interviews, documentation was noted as playing a significant role in
the development process, specifically with on-boarding new developers. A goal of
documentation is to lower the entry barrier for these new contributors. The
documentation provides information on how to get started, orients the user to
artifacts and the source code, and explains how the system works, including the
so-called simulation engine and interface.  This emphasis on documentation,
especially for new developers, is echoed in research software guidelines.
Multiple guidelines recommend a document explaining how to contribute to a
project, often named CONTRIBUTING \citep{Yo2021, BrettEtAl2021, WilsonEtAl2016,
ThielEtAl2020, vanGompelEtAl2016, OrvizEtAl2017, FLOSS2022, JimenezEtAl2017}.
Tutorials \citep{ThielEtAl2020}, trouble shooting guides \citep{OrvizEtAl2017,
SSI2022} and quick start examples \citep{ThielEtAl2020, vanGompelEtAl2016} are
also recommended.  \citep{SmithAndRoscoe2018} suggests including instructions
specifically for on-boarding new developers. For open source software in
general, \citep{Fogel2005} recommends providing tutorial style examples,
developer guidelines, demos and screenshots.

@@

From Section~\ref{Sec_CompareTools}: ``Interviews with the developers revealed a
potentially more frequent use of both unit testing and continuous integration,
compared to what was observed from studying the repository artifacts.''
